Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2025-04-23 review.qmd
Finalizing for pub

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  latex: default
  docx: 
    omitFigures: false
  pdf: default
    #keep-tex: true

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high signal to noise ratio (SNR), and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

>>>>

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  #latex: default
  docx: 
    omitFigures: true
  pdf: default
    #keep-tex: true

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high signal to noise ratio (SNR), and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

### 2025-04-23 review.tex
Finalizing for pub

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high signal to noise ratio (SNR), and they allow us to
analyze interactions between stimulus components using a convenient
mathematical framework. Here we describe how SSVEPs have been used to
study visual contrast over the past 70 years (Dawson, 1954). Because our
thinking about SSVEPs is well-described by simple mathematical models,
we embed code that illustrates key steps in the modelling and analysis.
This paper can therefore be used both as a review of the use of SSVEP in
measuring human contrast processing, and as an interactive learning aid.

@@

\bibitem[\citeproctext]{ref-Dawson1954}
Dawson GD (1954). \href{https://doi.org/10.1016/0013-4694(54)90007-3}{A
summation technique for the detection of small evoked potentials}.
\emph{Electroencephalography and Clinical Neurophysiology} \textbf{6,}
65--84.

>>>>

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high signal to noise ratio (SNR), and they allow us to
analyze interactions between stimulus components using a convenient
mathematical framework. Here we describe how SSVEPs have been used to
study visual contrast over the past 70 years. Because our thinking about
SSVEPs is well-described by simple mathematical models, we embed code
that illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@
